---
title: "Fonts"
author: "Carole Voulgaris"
date: "9/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(tidyverse)
library(extrafont)
library(extrafontdb)
```

The default fonts in R might not be that interesting to you. Here is how you can get a little more flexibility.

Let's start with a simple scatterplot.

```{r}
example_plot <- ggplot(trees, aes(x = Girth, 
                                  y = Height,
                                  color = Volume)) +
    geom_point() +
    ggtitle("Black Cherry Trees") +
    scale_x_continuous(name = "Tree diameter (inches)") +
    scale_y_continuous(name = "Tree height (feet)") +
    scale_color_continuous(name = "Volume of timber\n(cubic feet)")

example_plot
```

The default font is Arial or Helvetica, and it's fine. But maybe you want more options.

To make all the fonts on your computer available to R, type `font_import()` into your console and press Enter. You only need to ever do this once. It takes some time to import all the fonts, so you'll get a prompt to type y or n (yes or no) to confirm that you really want to do that. Type `y` and press Enter.

Now you can see a list of all the fonts that are available to you. Here's what I have on my computer (it's a PC).

```{r}
fonts()
```

```{r, message=FALSE}
loadfonts(device = "win")
```

To use one of these fonts in your plot, use the `theme()` function.

```{r}
example_plot +
  theme(text = element_text(family = "Garamond"))
```

Maybe you want to use different fonts for different text elements. No problem.

```{r}
example_plot +
  theme(title = element_text(family = "Copperplate Gothic Bold" ),
        axis.title = element_text(family = "Copperplate Gothic Light"),
        axis.text = element_text(family = "Calibri" ),
        legend.title = element_text(family = "Calibri", face = "bold"),
        legend.text = element_text(family = "Calibri" ))
```