
# Interactive maps

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(leaflet)
library(htmltools)
library(htmlwidgets)
library(rnaturalearth)
```

In this tutorial, I will download the world countries, color them based on their economic classification (column "economy" in the dataframe world), and show name, population and GDP per capita.

```{r}
world <- ne_countries(returnclass = "sf") %>% filter(name != "Antarctica")
```

Here you can find available basemaps in leaflet. Some of them will not work in R http://leaflet-extras.github.io/leaflet-providers/preview/.

This is how you can color polygons based on a specific value, similar to fill in ggplot.

```{r, warning=FALSE}
# Here, I'll create a color palette. This website gives a detailed tutorial on how to do that, for categorical and continuous variables.  https://rstudio.github.io/leaflet/colors.html

color_bins <- colorFactor(heat.colors(7), world$economy)

world_map1 <- leaflet(world) %>%
  addProviderTiles("Stamen.TonerLite") %>%
  addPolygons(fillColor = color_bins(world$economy), stroke = FALSE)

world_map1
```

### Highlights, dumb labels and popups

```{r, warning=FALSE}
# If you add a string in label or popup, it will display that same string for every polygon.

world_map2 <- leaflet(world) %>%
  addProviderTiles("Stamen.TonerLite") %>%
  addPolygons(fillColor = color_bins(world$economy), stroke = FALSE,
              highlightOptions = highlightOptions(fillColor = "red", fillOpacity = 0.5),
              label = "This is a country",
              popup = "That's all I know")
world_map2
```

### Smart labels and popups 


```{r, warning=FALSE}
# If you want to display info based on the values of different columns, you can simply call that column with label = dataframe$column
world_map3 <- leaflet(world) %>%
  addProviderTiles("Stamen.TonerLite") %>%
  addPolygons(fillColor = color_bins(world$economy), stroke = FALSE,
              highlightOptions = highlightOptions(fillColor = "red", fillOpacity = 0.5),
              label = world$name,
              popup = paste("Country:", world$name, "<br/>",
                            "Population:", world$pop_est/1000000,"M", "<br/>",
                            "GDP per capita: $", world$gdp_md_est/world$pop_est*1000000))
                           
world_map3
```


### Same, but with pretty numbers in popups

```{r, warning=FALSE}

world_map3 <- leaflet(world) %>%
  addProviderTiles("Stamen.TonerLite") %>%
  addPolygons(fillColor = color_bins(world$economy), stroke = FALSE,
              highlightOptions = highlightOptions(fillColor = "red", fillOpacity = 0.5),
              label = world$name,
              popup = paste("Country:", world$name, "<br/>",
                            "Population:", round(world$pop_est/1000000,1),"M", "<br/>",
                            "GDP per capita: $", format(round(world$gdp_md_est/world$pop_est*1000000),big.mark=",")))
                           
world_map3
```


### Map bounds, zoom controls and legend

```{r, warning=FALSE}
# Sometimes it makes sense to limit the extent of the map, especially if your region is small (a city). Also, it's a good practice to limit zooming in and out. 

world_map4 <- leaflet(world,
                      options = leafletOptions(minZoom = 2, maxZoom = 4)) %>%
  addProviderTiles("Stamen.TonerLite") %>%
  addPolygons(fillColor = color_bins(world$economy), stroke = FALSE,
              highlightOptions = highlightOptions(fillColor = "red", 
                                                  fillOpacity = 0.5),
              label = world$name,
              popup = paste("Country:", world$name, "<br/>",
                            "Population:", round(world$pop_est/1000000,1),"M", "<br/>",
                            "GDP per capita: $", format(round(world$gdp_md_est/world$pop_est*1000000),big.mark=","),  "<br/>")) %>%
  setMaxBounds( lng1 = -170,
                lat1 = -65,
                lng2 = 180,
                lat2 = 80) %>%
  addLegend(pal = color_bins,
             values = world$economy,
             title = "Economic Development",
            opacity = 1)

world_map4
```


```{r}
saveWidget(world_map4, file = "world_map4.html")
```


