---
title: "Week 2: Relationships and distributions"
author: "Carole Voulgaris"
date: "9/9/2020"
output: 
  html_document:
    theme: architect
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Fun with ggthemes!

``` {r load libraries, message=FALSE}
library(ggplot2)
library(ggthemes)
library(tidyverse)
library(tidycensus)
library(prettydoc)
```



```{r, message = FALSE, echo=FALSE}
vars <- c(tot_hh = "H001001",
         tot_rural_hh = "H002005",
         tot_rent_hh = "H004004",
         avg_hh_size = "H012001",
         tot_pop = "P003001",
         nh_white_pop = "P005003",
         nh_black_pop = "P005004",
         nh_asian_pop = "P005006",
         hispanic_pop = "P005010",
         med_age = "P013001")

SD_tracts <- get_decennial(geography = "tract", variables = vars,
                           state = "CA", county = "San Diego",
                           output = "wide") %>%
  mutate(rural = (tot_rural_hh == tot_hh) ,
         pct_rental = tot_rent_hh / tot_hh,
         majority_race_eth = case_when(
           nh_white_pop / tot_pop > 0.5 ~ "nh_white",
           nh_black_pop / tot_pop > 0.5 ~ "nh_black",
           nh_asian_pop / tot_pop > 0.5 ~ "nh_asian",
           hispanic_pop / tot_pop > 0.5 ~ "hispanic",
           (nh_white_pop + nh_black_pop + nh_asian_pop + hispanic_pop) /
             tot_pop < 0.5 ~ "other",
           TRUE ~ "None")) %>%
  filter(tot_pop > 0, tot_hh > 0) %>%
  select(GEOID, rural, majority_race_eth, 
         pct_rental, avg_hh_size, tot_pop, med_age) 
```


Here's a figure from the assignment tutorial:

```{r}
figure <- ggplot(SD_tracts, 
                 aes(x = pct_rental,
                     y = med_age,
                     color = majority_race_eth)) +
  geom_point() +
  scale_x_continuous(name = "Percent of households in rented homes",
                     breaks = seq(0, 1, by = 0.1),
                     labels = paste(seq(0, 100, by = 10), "%", sep = "")) +
  scale_y_continuous(name = "Median age within tract",
                     breaks = seq(10, 80, by = 10)) +
  scale_color_discrete(name = "Majority race",
                       labels = c("Hispanic",
                                  "Non-Hispanic Asian",
                                  "Non-Hispanic white",
                                  "No majority")) 

figure
```

We can use the theme function to adjust the appearance of the figure:

```{r}
figure +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_line(colour = "black",
                                  linetype = "dashed"),
        legend.key = element_rect(fill = "white"))
```

Or we can use one of many pre-defined themes from either the `ggplot2` or the `ggthemes` package.

Here is a theme that follows the style of the fivethirtyeight website:

```{r}
figure +
  theme_fivethirtyeight()
```

And here's one that follows the style of The Economist (the magazine):

```{r}
figure +
  theme_economist()
```

If you creating graphs in Excel, you might try this:

```{r}
figure +
  theme_excel()
```

Here is a theme that follows the guidelines of Edward Tufte:

```{r}
figure +
  theme_tufte()
```

Here is a theme that follows the guidelines of Stephen Few:

```{r}
figure +
  theme_few()
```


I can also take a pre-defined theme and customize it further:

```{r}
figure +
  theme_few() +
  theme(legend.position = "bottom")
```
