---
title: "Week 2, Wednesday Lecture Notes"
author: "Carole Voulgaris"
date: "9/9/2020"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)

data <- read_csv("dataset_vis.csv")
time <- read_csv("time_data.csv")
```

## Sample dataset

Here are the first few lines of a person-level dataset with five variables:

* Gender
* Employment status
* Income
* Population density within a mile of home
* Daily vehicle miles traveled

***

``` {r, echo=FALSE}
knitr::kable(data[1:7, ])
```
## Visualizing a continuous distribution

Here are four alternative ways to visualize the distribution of income within the dataset. How might you interpret these differently?

```{r, echo = FALSE}
ggplot(data, aes(x = income)) +
  geom_histogram(bins = 50, color = "gray") +
  scale_x_continuous(name = "Annual household income",
                     breaks = inc_breaks <- seq(0,350000, by = 50000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ","))) +
  scale_y_continuous(name = "Number of people") +
  theme_bw()
```


```{r, echo = FALSE}
ggplot(data, aes(y = income)) +
  geom_boxplot() +
  scale_y_continuous(name = "Annual household income",
                     breaks = inc_breaks <- seq(0,350000, by = 50000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ","))) +
  theme_bw() +
  theme(axis.text.x = element_blank())
```

```{r, echo = FALSE}
ggplot(data, aes(x = income)) +
  geom_freqpoly(bins = 100) +
  scale_x_continuous(name = "Annual household income",
                     breaks = inc_breaks <- seq(0,350000, by = 50000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ","))) +
  theme_bw() 
```

```{r, echo = FALSE}
ggplot(data, aes(x = 1, y = income)) +
  geom_violin() +
  scale_y_continuous(name = "Annual household income",
                     breaks = inc_breaks <- seq(0,350000, by = 50000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ","))) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.title.x.bottom = element_blank())
```
```{r, echo = FALSE}
ggplot(data, aes(x = 1, y = income)) +
  geom_jitter(alpha = 0.2, size = 0.3) +
  scale_y_continuous(name = "Annual household income",
                     breaks = inc_breaks <- seq(0,350000, by = 50000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ","))) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.title.x.bottom = element_blank())
```

## Distributions of categorical variables

Here are three ways to visualize the distribution of a categorical variable. How else could you visualize these distributions? Why might you choose one option or another? 

```{r, echo = FALSE}
ggplot(data, aes(x = 1, fill = employment)) +
  geom_bar(position = "fill") +
  scale_y_continuous(name = "Percent of population",
                     breaks = seq(0, 1, by = 0.1),
                     labels = paste(seq(0,100, by = 10), "%", sep = "")) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.title.x.bottom = element_blank())
```
```{r, echo = FALSE}
ggplot(data, aes(x = 1, fill = employment)) +
  geom_bar(position = "fill") +
  scale_y_continuous(name = "Percent of population",
                     breaks = seq(0, 1, by = 0.1),
                     labels = paste(seq(0,100, by = 10), "%", sep = "")) +
  theme_void() +
  coord_polar(theta = "y") 
```
```{r, echo = FALSE}
ggplot(data, aes(x = employment)) +
  geom_bar() +
  scale_x_discrete(name = "Employment status") +
  scale_y_continuous(name = "Number of people") +
  theme_bw() 
```

## Relationships between continuous variables

Here are some alternative representations of the relationship between VMT and income.

```{r, echo = FALSE}
ggplot(data, aes(x = VMT, y = income)) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_x_continuous(name = "Daily vehicle miles traveled") +
  scale_y_continuous(name = "Annual household income",
                     breaks = inc_breaks <- seq(0,350000, by = 50000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ","))) +
  theme_bw() 
```
```{r, echo = FALSE}
ggplot(data, aes(x = VMT, y = income)) +
  geom_line() +
  scale_x_continuous(name = "Daily vehicle miles traveled") +
  scale_y_continuous(name = "Annual household income",
                     breaks = inc_breaks <- seq(0,350000, by = 50000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ","))) +
  theme_bw() 
```

```{r, echo = FALSE}
ggplot(data, aes(x = VMT, y = income)) +
  geom_point(alpha = 0.1, size = 0.3, color = "gray") +
  stat_smooth(color = "black", method = 'gam', formula = y ~ s(x, bs = "cs")) +
  scale_x_continuous(name = "Daily vehicle miles traveled") +
  scale_y_continuous(name = "Annual household income",
                     breaks = inc_breaks <- seq(0,350000, by = 50000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ","))) +
  theme_bw() 
```

```{r, echo = FALSE}
ggplot(data, aes(x = VMT, y = income)) +
  geom_point(alpha = 0.1, size = 0.3, color = "gray") +
  stat_smooth(color = "black", method = 'lm', formula = y ~ x) +
  scale_x_continuous(name = "Daily vehicle miles traveled") +
  scale_y_continuous(name = "Annual household income",
                     breaks = inc_breaks <- seq(0,350000, by = 50000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ","))) +
  theme_bw() 
```

```{r, echo = FALSE}
ggplot(data, aes(x = income, y = VMT)) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_y_continuous(name = "Daily vehicle miles traveled") +
  scale_x_continuous(name = "Annual household income",
                     breaks = inc_breaks <- seq(0,350000, by = 50000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ","))) +
  theme_bw() 
```

```{r, echo = FALSE}
ggplot(data, aes(x = income, y = VMT)) +
  geom_line(size = 0.1) +
  scale_y_continuous(name = "Daily vehicle miles traveled") +
  scale_x_continuous(name = "Annual household income",
                     breaks = inc_breaks <- seq(0,350000, by = 50000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ","))) +
  theme_bw() 
```

## Log scales and linear scale

All of the above example show income on a linear scale. What if we showed it on a log scale?

```{r, echo = FALSE}
ggplot(data, aes(x = income)) +
  geom_histogram(bins = 50, color = "gray") +
  scale_x_continuous(name = "Annual household income",
                     breaks = inc_breaks <- seq(0,350000, by = 50000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ","))) +
  scale_y_continuous(name = "Number of people") +
  theme_bw()
```
```{r, echo = FALSE}
ggplot(data, aes(x = income)) +
  geom_histogram(bins = 50, color = "gray") +
  scale_x_continuous(name = "Annual household income",
                     breaks = inc_breaks <- c(2000, 5000, 10000, 
                                              20000, 50000, 100000,
                                              200000, 500000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ",")),
                     trans = "log") +
  scale_y_continuous(name = "Number of people") +
  theme_bw()
```

```{r, echo = FALSE}
ggplot(data, aes(x = income, y = VMT)) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_y_continuous(name = "Daily vehicle miles traveled") +
  scale_x_continuous(name = "Annual household income",
                     breaks = inc_breaks <- seq(0,350000, by = 50000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ","))) +
  theme_bw() 
```

```{r, echo = FALSE}
ggplot(data, aes(x = income, y = VMT)) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_y_continuous(name = "Daily vehicle miles traveled") +
  scale_x_continuous(name = "Annual household income",
                     breaks = inc_breaks <- c(2000, 5000, 10000, 
                                              20000, 50000, 100000,
                                              200000, 500000),
                     labels = paste("$", 
                                    formatC(inc_breaks, 
                                            format = "d",
                                              big.mark = ",")),
                     trans = "log") +
  theme_bw() 
```

## Cartesian or Polar Coordinates?

Let's look at some time-series data. Here are selected rows from a day-level dataset with five variables:

* Number of day since we started measuring total vehicle miles traveled
* Day of the week
* Day of the week (as a number)
* Week number
* Million Vehicle Miles Traveled in the city that day

***

``` {r, echo=FALSE}
knitr::kable(time[13:26, ])
```
Here are some graphs that shows how MVMT has changed over time.

```{r, echo = FALSE}
ggplot(time, aes(x = day_count, y = MVMT)) +
  geom_point() +
  scale_x_continuous(name = "Number of days since counts began",
                     breaks= seq(0, 60, by = 5)) +
  theme_bw()
```
```{r, echo = FALSE}
ggplot(time, aes(x = day_count, y = MVMT)) +
  geom_line() +
  scale_x_continuous(name = "Number of days since counts began",
                     breaks= seq(0, 60, by = 5)) +
  theme_bw()
```
```{r, echo = FALSE}
ggplot(time, aes(x = day_num, y = MVMT, color = as.character(Week))) +
  geom_line() +
  scale_x_continuous(name = "Day of the week",
                     breaks = seq(1, 7, by=1),  
                     labels = c("Sunday",
                                "Monday",
                                "Tuesday",
                                "Wednesday",
                                "Thursday",
                                "Friday",
                                "Saturday")) +
  scale_color_viridis_d(name = "Week number") +
  theme_bw()
```

A polar coordinate system is useful for showing cycles:

```{r, echo = FALSE}
ggplot(time, aes(x = day_num, y = MVMT, color = as.character(Week))) +
  geom_line() +
  scale_x_continuous(name = "Day of the week",
                     breaks = seq(1, 7, by=1),  
                     labels = c("Sunday",
                                "Monday",
                                "Tuesday",
                                "Wednesday",
                                "Thursday",
                                "Friday",
                                "Saturday")) +
  scale_color_viridis_d(name = "Week number") +
  coord_polar() +
  theme_bw()
```

(Polar coodinates are also useful for pie charts)

